The impact of gas bulk rotation on the Lyman $\alpha$ line

### 2014-05-22 paper/ms.tex
conclusions

\cite{Verhamme12} studied \lya using two high resolution simulations
of individual galaxies. The main purpose of the study was to asses the
impact of two different InterStellar Medium (ISM) prescriptions. However, each simulated
galaxy had a disc structure with a clear rotation pattern in the ISM
and inflowing gas int the circum-galactic region. The configuration
had an axial symmetry and they reported a strong dependence of the
escape fraction and the total line intensity as a function of the
$\theta$ angle. In our results, none of these two quantities has a
dependence either on the inclination angle or the rotational
velocity. Therefore, the effect reported by \cite{Verhamme} is
consistent to the different hydrogen optical depth for different
viewing angles and not as an effect of the bulk rotation.

The presence of single peaked profiles has been associated to
inflow/outflow dynamics \citep{Verhamme06,DijkstraKramer}. In this
paper we show that gas bulk rotation can also be considered as a
probable origin for that behavior, provided that the observed single peak is highly
symmetric. Similarly, in the case of double peaked lines with a high
level of flux at the line center, rotation also deserves to be considered in
the pool of possible bulk flows responsible for that feature,
specially if the two peaks have similar intensities. The case of
triple peaked lines is another clear feature of gas rotation under the
additional condition of an off-centered emission.

@@

This paper quantifies for the first time in the literature the effects
of rotation in the morphology of the \ly emission line in star forming
galaxies.  The results are based on the study of an homogeneous sphere
of gas with solid body rotation. We explore a range of models by varying
the rotational speed, hydrogen optical depth, dust optical depth and
initial distribution of \ly photons with respect to the gas
density. As a cross-validation, we obtain our results from two
independently developed Monte-Carlo radiative transfer codes.

Our main result is that rotation clearly impacts the \ly line
morphology. The morphology of the observed spectra for observers at different
viewing angle $\theta$ changes while the flux remains constant for all $\theta$,
for $\mu \sim 1$ double peaked lines can make transitions into single
peaked lines as the rotational velocity increases.

@@

\item Rotation induce a considerable asymmetry in the morphology 
 of the spectra when the viewing angle $\theta$. This occurs for
 all of our models, in particular models with $\mu\sim 1$ i.e for 
 observes aligned with the axis of rotation the rotation does not 
 produce a significance effect while for $\mu\sim 0$ the effect in 
 the FWHM, and Maxima is significant and the number of scatterings.

%\item In spite of the asymmetry induced by rotation the spatial
%  anisotropy in the emission is usually below the
%  $10\%$ level. This justifies our consideration of angle-averaged
%  effects to quantify the emission line of rotating clouds.

\item For $\mu \sim 0$ The line width increases with rotational velocity. The line width
  approximately follows the functional for  ${\rm FWHM}^2 = {\rm FWHM}_{
    0}^2 + (V_{\rm max}/\lambda)^2$ where FWHM$_{0}$ indicates the line
  width for the static case and $\lambda$ is a constant determined from
  the radiative transfer results it is $\lambda_{\rm c}=0.84 \pm 0.06$ and
  $\lambda_{\rm h}=0.54\pm 0.1$ for the central and homogeneous source
  distribution, respectively.

\item A single peaked line emerges for high rotational velocities in
  the case of homogeneously distributed sources. These cases occur when
  the rotational velocity is larger than half FWHM$_0$ and $\mu\sim 0$.

\item For central sources we find that the number of scatterings
does not significantly decrease with rotation. This is due to the 
fact that the vast majority of scatterings events are resonant, 
during which the mean free path is very short and the effect of gas
bulk rotation is not enough to affect \ly photons. In consequence, 
the escape fraction neither the position of the peaks depends on the
rotational velocity. However, there is a flux increase with rotational
velocity in the central parts of the line.

\item For homogeneous sources we find that the changes in the
line morphology are linked to the reduction in the average number of
scatterings. In the second part of \S \ref{sec:widthpeak} we show
that, as rotational velocity increases, a large fraction of photons
escape with less than $10$ scatterings. The average number of
scatterings decreases a $40\%$ compared to the static case. 
This reduction in the number of scatterings decreases the probability
for a photon to move away from the line's center. This is also
translated into an increase of the escape fraction by a factor of
$\sim 2$ with respect to the static situation.

@@

%Comparing our results with recent observed LAEs we find that many 
%morphological features such as high central line flux, single peak
%profiles  could be explained by gas bulk  rotation present in these
%LAEs.

>>>>

\cite{Verhamme12} studied \ly using two high resolution simulations
of individual galaxies. The main purpose of the study was to asses the
impact of two different InterStellar Medium (ISM) prescriptions. However, each simulated
galaxy had a disc structure with a clear rotation pattern in the ISM
and inflowing gas int the circum-galactic region. The configuration
had an axial symmetry and they reported a strong dependence of the
escape fraction and the total line intensity as a function of the
$\theta$ angle. In our results, none of these two quantities has a
dependence either on the inclination angle or the rotational
velocity. Therefore, the effect reported by \cite{Verhamme12} is
consistent with being a consequence of the different hydrogen optical
depth for different viewing angles and not as an effect of the bulk
rotation.

The presence of single peaked profiles has been associated to
inflow/outflow dynamics \citep{Verhamme06,DijkstraKramer}. In this
paper we show that gas bulk rotation can also be considered as a
probable origin for that behavior, provided that the observed single peak is highly
symmetric. Similarly, in the case of double peaked lines with a high
level of flux at the line center, rotation also deserves to be considered in
the pool of possible bulk flows responsible for that feature,
specially if the two peaks have similar intensities. This highlights
that in order to interpret the observations of \ly emitting galaxies
it is necessary to consider the possible effect of different
inclination angles along the line of sight.

@@

This paper quantifies for the first time in the literature the effects
of gas rotation in the morphology of the \ly emission line in star forming
galaxies.  The results are based on the study of an homogeneous sphere
of gas with solid body rotation. We explore a range of models by varying
the rotational speed, hydrogen optical depth, dust optical depth and
initial distribution of \ly photons with respect to the gas
density. As a cross-validation, we obtain our results from two
independently developed Monte-Carlo radiative transfer codes.

Our main result is two-fold. First,rotation clearly impacts the \ly line
morphology. Second, rotation introduces an anisotropy for different
viewing angles. For viewing angles close to the poles the line is
double peaked and it makes a transition to a single peaked line for
high rotational velocities and viewing angles along the equator. This
trend is stronger for spheres with homogeneusly distributed radiation
sources.  In contrast, the total flux in the line remains independent
of the viewing angle.

The escape fraction remains constant for diffenret viewing angles only
in the case of centrally distributed sources. In contrast, for an homogeneous
distribution the escape fraction decreases with high rotation 
fraction remain constant for different viewing angles. We showed that
these results can be interpreted in terms of the average number of
scatterings for the photons, a quantity that also remains constant
independent of the viewing angle and the rotational velocity for
centrally distributed sources and decreases for homogenously
distributed source.

@@

\item Rotation induces a considerable asymmetry in the morphology 
 of the spectra when the viewing angle $\theta$. This occurs for
 all of our models. The asymmetry is such that observeds with
 $\cos\theta\equiv \mu\sim 1$, i.e. observes aligned with the rotation
 axis, the line is double peaked while for $\mu\sim 0$ and high
 rotation veolcities the linebecomes double peaked.

\item For $\mu \sim 0$ The line width increases with rotational velocity. The line width
  approximately follows the functional for  ${\rm FWHM}^2 = {\rm FWHM}_{
    0}^2 + (V_{\rm max}/\lambda)^2$ where FWHM$_{0}$ indicates the line
  width for the static case and $\lambda$ is a constant determined from
  the radiative transfer results it is $\lambda_{\rm c}=0.84 \pm 0.06$ and
  $\lambda_{\rm h}=0.54\pm 0.10$ for the central and homogeneous source
  distribution, respectively.

\item A single peaked line emerges for high rotational velocities in
  the case of homogeneously distributed sources. These cases occur when
  the rotational velocity is larger than half FWHM$_0$ and $\mu\sim 1$.

\item 
For central sources we find that the escape fraction does not change
with rotational velocity nor with the viewing angle. We interpret this
in terms of the average number of scatterings which does not
significantly changes with rotation. This is due to the  
  fact that the vast majority of scatterings events are resonant, 
  during which the mean free path is very short and the effect of gas
  bulk rotation is not enough to affect \ly photons.

\item For homogeneous sources we find that the  as rotational velocity
  increases, the escape fraction rises by a factor of $\sim 2$ with
  respect to the static situation. We also interpret this change in
  terms of the average number of scatterings. In the second part of \S
  \ref{sec:widthpeak} we show that the average number of scatterings
  decreases a $40\%$ compared to the static case, while a large
  fraction of photons escape with less than $10$ scatterings.

@@

Comparing our results with recent observed LAEs we find that many 
morphological features such as high central line flux, single peak
profiles  could be explained by gas bulk  rotation present in these
LAEs.
